---
title: "Unsupervised Learning with Text Data"
date: 2019-03-07T13:30:00  # Schedule page publish date.
draft: false
type: "talk"

output:
  blogdown::html_page:
    toc: false
    number_sections: false

# Talk start and end times.
#   End time can optionally be hidden by prefixing the line with `#`.
time_start: 2019-03-11T13:30:00
time_end: 2019-03-11T15:00:00

# Authors. Comma separated list, e.g. `["Bob Smith", "David Jones"]`.
authors: []

# Abstract and optional shortened version.
abstract: ""
abstract_short: "Introduction to text data structures, basic visualizations, dimension reduction, and topic modeling."

# Location of event.
location: "Room 247, Saieh Hall for Economics, Chicago, IL"

# Is this a selected talk? (true/false)
selected: false

# Tags (optional).
#   Set `tags: []` for no tags, or use the form `tags: ["A Tag", "Another Tag"]` for one or more tags.
tags: []

# Links (optional).
url_pdf: ""
url_slides: ""
url_video: ""
url_code: ""

# Does the content use math formatting?
math: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Overview

* Identify the basic workflow for conducting text analysis
* Define sentiment analysis and review its use in academic research
* Implement latent semantic analysis using text data
* Introduce Latent Dirchlet allocation and apply it to example data

## Before class

## Slides and class materials

```{r slides}
knitr::include_url("/slides/unsupervised-text.html", height = "505px")
```

* [Latent semantic analysis](/notes/latent-semantic-analysis/)
* [Latent Dirchlet allocation](/notes/latent-dirchlet-allocation/)

### Additional readings

* [Text Mining with R: A Tidy Approach](https://www.tidytextmining.com/)
* [Natural Language Toolkit](https://www.nltk.org/)

## What you need to do

