---
title: "Best practices in designing statistical graphics"
author: "[MACS 30200](https://github.com/css-research/course/) <br /> University of Chicago"
output: rcfss::xaringan
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(cache = TRUE, echo = TRUE, message = FALSE, warning = FALSE,
                      fig.retina = 2, fig.width = 12)

library(tidyverse)
library(ggthemes)
library(knitr)
library(broom)
library(stringr)

options(digits = 3)
set.seed(1234)
theme_set(theme_minimal(base_size = rcfss::base_size))
```

# Tufte's world

* Core purpose of visualization is to communicate quantitative information
    * Art is secondary
    * "Above all else show the data"

--

$$\max(\text{Data-ink ratio}) = \frac{\text{data-ink}}{\text{total ink used to print the graphic}}$$

* What should we consider to be part of the "data-ink"?

---

# What is integral?

```{r integral}
p <- ggplot(mpg, aes(cty, hwy)) +
  geom_point()
p
```

---

# Remove the non-data elements

```{r integral-void}
p + theme_void()
```

---

# Remove background color

```{r integral-background}
p + theme_bw()
```

---

# Remove panel box

```{r integral-panelbox}
p + theme_bw() +
    theme(panel.background = element_blank(),
          panel.border = element_blank())
```

---

# Remove minor grid lines

```{r integral-minor}
p + theme_bw() +
    theme(panel.background = element_blank(),
          panel.border = element_blank(),
          strip.background = element_blank(),
          plot.background = element_blank(),
          axis.line = element_blank(),
          panel.grid.minor = element_blank())
```

---

# Remove all grid lines

```{r integral-major}
p + theme_bw() +
    theme(panel.background = element_blank(),
          panel.border = element_blank(),
          strip.background = element_blank(),
          plot.background = element_blank(),
          axis.line = element_blank(),
          panel.grid = element_blank())
```

---

# Remove tick marks

```{r integral-tick}
p + theme_bw() +
    theme(panel.background = element_blank(),
          panel.border = element_blank(),
          strip.background = element_blank(),
          plot.background = element_blank(),
          axis.line = element_blank(),
          panel.grid = element_blank(),
          axis.ticks = element_blank())
```

---

# Use serif font

```{r integral-serif}
p + theme_bw(base_family = "serif") +
    theme(panel.background = element_blank(),
          panel.border = element_blank(),
          strip.background = element_blank(),
          plot.background = element_blank(),
          axis.line = element_blank(),
          panel.grid = element_blank(),
          axis.ticks = element_blank())
```

---

# Chart junk

* Vibrating moire effects
* The grid
* The duck

---

# Line plot

```{r line}
x <- 1967:1977
y <- c(0.5, 1.8, 4.6, 5.3, 5.3, 5.7, 5.4, 5, 5.5, 6, 5)
d <- tibble(x, y)

ggplot(d, aes(x, y)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(breaks = seq(1, 6, 1), label = sprintf("$%s", seq(300, 400, 20))) +
  labs(title = "Per capita budget expandures",
       x = "Year",
       y = "Per capita budget expandures\nin constant dollars")
```

---

# Minimal line plot

```{r minline}
ggplot(d, aes(x, y)) +
  geom_line() +
  geom_point(size = 3) +
  theme_tufte(base_size = 15) +
  theme(axis.title = element_blank()) +
  geom_hline(yintercept = c(5, 6), lty = 2) +
  scale_y_continuous(breaks = seq(1, 6, 1), label = sprintf("$%s", seq(300, 400, 20))) +
  scale_x_continuous(breaks = x, label = x) +
  annotate(
    "text",
    x = c(1977, 1977.2),
    y = c(1.5, 5.5),
    adj = 1,
    family = "serif",
    label = c("Per capita\nbudget expandures\nin constant dollars", "5%")
  )
```

---

# Boxplot

```{r boxplot}
ggplot(quakes, aes(factor(mag), stations)) +
  geom_boxplot() +
  labs(title = "Fiji earthquakes",
       x = "Richter Magnitude",
       y = "Number of stations reporting earthquakes")
```

---

# Minimal boxplot

```{r minboxplot}
ggplot(quakes, aes(factor(mag), stations)) +
  theme_tufte() +
  geom_tufteboxplot(outlier.colour = "transparent") +
  theme(axis.title = element_blank()) +
  annotate(
    "text",
    x = 8,
    y = 120,
    adj = 1,
    family = "serif",
    label = c(
      "Number of stations \nreporting Richter Magnitude\nof Fiji earthquakes (n=1000)"
    )
  )
```

---

# Minimal offset boxplot

```{r offset-boxplot}
ggplot(quakes, aes(factor(mag), stations)) +
  theme_tufte() +
  geom_tufteboxplot(median.type = "line") +
  theme(axis.title = element_blank()) +
  annotate(
    "text",
    x = 8,
    y = 120,
    adj = 1,
    family = "serif",
    label = c(
      "Number of stations \nreporting Richter Magnitude\nof Fiji earthquakes (n=1000)"
    )
  )
```

---

# Barchart

```{r bar}
library(psych)
library(reshape2)

d <- melt(colMeans(msq[, c(2, 7, 34, 36, 42, 43, 46, 55, 68)], na.rm = T) *
            10)
d$trait <- rownames(d)

p <- ggplot(d, aes(x = trait, y = value)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(breaks = seq(1, 5, 1)) +
  labs(title = "Watson et al., 1998",
       subtitle = "N = 3896",
       x = "Negative emotion traits",
       y = "Average score")
p
```

---

# Minimal barchart

```{r minbar}
ggplot(d, aes(x = trait, y = value)) +
  theme_tufte(base_size = 14, ticks = F) +
  geom_bar(width = 0.25, fill = "gray", stat = "identity") +
  theme(axis.title = element_blank()) +
  scale_y_continuous(breaks = seq(1, 5, 1)) +
  geom_hline(yintercept = seq(1, 5, 1),
             col = "white",
             lwd = 1) +
  annotate(
    "text",
    x = 3.5,
    y = 5,
    adj = 1,
    family = "serif",
    label = c(
      "Average scores\non negative emotion traits
from 3896 participants\n(Watson et al., 1988)"
    )
  )
```

---

# When is redundancy better?

![Traditional bar chart of crime in the city of San Francisco, 2009-10. Source: [Visualizing Time with the Double-Time Bar Chart](http://dougmccune.com/blog/2011/04/26/visualizing-time-with-the-double-time-bar-chart/)](http://dougmccune.com/blog/wp-content/uploads/2011/04/sk_time_chart1.png)

![Double-time bar chart of crime in the city of San Francisco, 2009-10. Source: [Visualizing Time with the Double-Time Bar Chart](http://dougmccune.com/blog/2011/04/26/visualizing-time-with-the-double-time-bar-chart/)](http://dougmccune.com/blog/wp-content/uploads/2011/04/burglary1.png)
    
---

# Does minimalism really help?

![Chart from Harvard magazine. Source: [Involuntary head-shaking is probably not an intended consequence of data visualization](http://junkcharts.typepad.com/junk_charts/2016/11/involuntary-head-shaking-is-probably-not-an-intended-consequence-of-data-visualization.html)](http://junkcharts.typepad.com/.a/6a00d8341e992c53ef01b7c8b60cd9970b-pi)

---

# Redesigned chart

![Redesigned chart from Harvard magazine. Source: [Involuntary head-shaking is probably not an intended consequence of data visualization](http://junkcharts.typepad.com/junk_charts/2016/11/involuntary-head-shaking-is-probably-not-an-intended-consequence-of-data-visualization.html)](http://junkcharts.typepad.com/.a/6a00d8341e992c53ef01b8d23fe1c7970c-pi)

---

# Visualizing statistical models

* Do it right
* Present findings in substantive terms
* Show degree of confidence
* Show your data when you can

---

# Visualizing statistical models

## Tables

* Parameter estimates
* Standard errors
* Model statistics

--

## Visualizations

* Predicted values
* Original data points

---

# Tables vs. graphs

* [Charts or tables?](https://www.jstor.org/stable/20446574)


