---
title: "Resampling Methods"
date: 2019-01-24T13:30:00  # Schedule page publish date.
draft: false
type: "talk"

output:
  blogdown::html_page:
    toc: false
    number_sections: false

# Talk start and end times.
#   End time can optionally be hidden by prefixing the line with `#`.
time_start: 2019-01-28T13:30:00
time_end: 2019-01-28T15:00:00

# Authors. Comma separated list, e.g. `["Bob Smith", "David Jones"]`.
authors: []

# Abstract and optional shortened version.
abstract: ""
abstract_short: "Introduction to cross-validation and bootstrapping for model validation."

# Location of event.
location: "Room 247, Saieh Hall for Economics, Chicago, IL"

# Is this a selected talk? (true/false)
selected: false

# Tags (optional).
#   Set `tags: []` for no tags, or use the form `tags: ["A Tag", "Another Tag"]` for one or more tags.
tags: []

# Links (optional).
url_pdf: ""
url_slides: ""
url_video: ""
url_code: ""

# Does the content use math formatting?
math: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Overview

* Review the difference between the training and test MSE and the need for a test set of data
* Define resampling methods
* Compare and contrast the validation set approach with leave-one-out and $k$-fold cross-validation
* Define bootstrapping and explain when it can be used in research

## Before class

* ISL ch 5

## Slides and class materials

```{r slides}
knitr::include_url("/slides/resampling.html", height = "505px")
```

* 

### Additional readings


## What you need to do

