---
title: "Interpreting Machine Learning Models"
date: 2019-01-26T13:30:00  # Schedule page publish date.
draft: false
type: "talk"

output:
  blogdown::html_page:
    toc: false
    number_sections: false

# Talk start and end times.
#   End time can optionally be hidden by prefixing the line with `#`.
time_start: 2019-01-30T13:30:00
time_end: 2019-01-30T15:00:00

# Authors. Comma separated list, e.g. `["Bob Smith", "David Jones"]`.
authors: []

# Abstract and optional shortened version.
abstract: ""
abstract_short: "Overview of methods for interpreting machine learning models, including both global and local interpretations."

# Location of event.
location: "Room 247, Saieh Hall for Economics, Chicago, IL"

# Is this a selected talk? (true/false)
selected: false

# Tags (optional).
#   Set `tags: []` for no tags, or use the form `tags: ["A Tag", "Another Tag"]` for one or more tags.
tags: []

# Links (optional).
url_pdf: ""
url_slides: ""
url_video: ""
url_code: ""

# Does the content use math formatting?
math: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Overview

* Interpret regression results using marginal effects and `margins`
* Distinguish between global and local interpretations of machine learning models
* Define model agnostic metrics for global interpretations
* Define model agnostic metrics for local interpretations
* Implement metrics using `iml`/`scaler`, `DALEX`, and `lime`

## Before class

* ["Model-Agnostic Methods" in *Interpretable Machine Learning*](https://christophm.github.io/interpretable-ml-book/agnostic.html)

## Slides and class materials

```{r slides}
knitr::include_url("/slides/interpret-ml.html", height = "505px")
```

* 

### Additional readings


## What you need to do

