---
title: "More Classification Methods"
date: 2019-01-19T13:30:00  # Schedule page publish date.
draft: false
type: "talk"

output:
  blogdown::html_page:
    toc: false
    number_sections: false

# Talk start and end times.
#   End time can optionally be hidden by prefixing the line with `#`.
time_start: 2019-01-23T13:30:00
time_end: 2019-01-23T15:00:00

# Authors. Comma separated list, e.g. `["Bob Smith", "David Jones"]`.
authors: []

# Abstract and optional shortened version.
abstract: ""
abstract_short: "An overview of linear discriminant analysis, naive Bayes, and nearest-neighbors methods for classification."

# Location of event.
location: "Room 247, Saieh Hall for Economics, Chicago, IL"

# Is this a selected talk? (true/false)
selected: false

# Tags (optional).
#   Set `tags: []` for no tags, or use the form `tags: ["A Tag", "Another Tag"]` for one or more tags.
tags: []

# Links (optional).
url_pdf: ""
url_slides: ""
url_video: ""
url_code: ""

# Does the content use math formatting?
math: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Overview

* Define nearest neighbors regression/classification
* Demonstrate how to implement and interpret nearest neighbors algorithms
* Define Naive Bayes classifier and demonstrate its usefulness
* Explain linear discriminant analysis and distinguish from quadratic discriminant analysis

## Before class

* ISL ch 2.2.3, 4.4
* ESL ch 4.3, 13.3-4

## Slides and class materials

```{r slides}
knitr::include_url("/slides/lda-nearest-neighbor.html", height = "505px")
```

* [Nearest neighbors](/notes/nearest-neighbors/)
* [Naive Bayes](/notes/naive-bayes/)
* [Linear/quadratic discriminant analysis](/notes/linear-discriminant-analysis/)

### Additional readings


## What you need to do

* Start [homework 03](/homework/classification/)
