---
title: "Linear Regression"
date: 2019-01-10T13:30:00  # Schedule page publish date.
draft: false
type: "talk"

output:
  blogdown::html_page:
    toc: false
    number_sections: false

# Talk start and end times.
#   End time can optionally be hidden by prefixing the line with `#`.
time_start: 2019-01-14T13:30:00
time_end: 2019-01-14T15:00:00

# Authors. Comma separated list, e.g. `["Bob Smith", "David Jones"]`.
authors: []

# Abstract and optional shortened version.
abstract: ""
abstract_short: "An overview of linear regression."

# Location of event.
location: "Room 247, Saieh Hall for Economics, Chicago, IL"

# Is this a selected talk? (true/false)
selected: false

# Tags (optional).
#   Set `tags: []` for no tags, or use the form `tags: ["A Tag", "Another Tag"]` for one or more tags.
tags: []

# Links (optional).
url_pdf: ""
url_slides: ""
url_video: ""
url_code: ""

# Does the content use math formatting?
math: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Overview

* Define linear regression
* Explain why linear regression is so popular for statistical learning
* Identify the core assumptions of linear regression
* Demonstrate practical applications of linear regression

## Before class

* ISL ch 3
* ESL ch 3.2


## Slides and class materials

```{r slides}
knitr::include_url("/slides/select-model.html", height = "505px")
```

* [Statistical learning](/notes/stat-learning)
* [Assessing model accuracy](/notes/model-accuracy)

### Additional readings

* PRML ch 3

## What you need to do

* [Register your GitHub username for the course](https://goo.gl/forms/vqwIztpqLb5Dhs3e2) -- all homework assignments will be in **private repositories**. Private repos can only be seen and edited by members of our [course organization](https://github.com/css-model). Once you register your GitHub account, I will invite you to join the course organization. If you don't register your account, you won't have access to any of the homework assignments.
* Start [homework 01](/homework/theory-model-selection/)
