---
title: "Logistic Regression"
date: 2019-01-12T13:30:00  # Schedule page publish date.
draft: false
type: "talk"

output:
  blogdown::html_page:
    toc: false
    number_sections: false

# Talk start and end times.
#   End time can optionally be hidden by prefixing the line with `#`.
time_start: 2019-01-16T13:30:00
time_end: 2019-01-16T15:00:00

# Authors. Comma separated list, e.g. `["Bob Smith", "David Jones"]`.
authors: []

# Abstract and optional shortened version.
abstract: ""
abstract_short: "An overview of logistic regression."

# Location of event.
location: "Room 247, Saieh Hall for Economics, Chicago, IL"

# Is this a selected talk? (true/false)
selected: false

# Tags (optional).
#   Set `tags: []` for no tags, or use the form `tags: ["A Tag", "Another Tag"]` for one or more tags.
tags: []

# Links (optional).
url_pdf: ""
url_slides: ""
url_video: ""
url_code: ""

# Does the content use math formatting?
math: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Overview

* Define a classification problem
* Demonstrate why linear regression is a poor method to use for classification problems
* Introduce the logistic regression functional form
* Distinguish between probability, odds, and log-odds
* Identify how to calculate predicted probabilities
* Implement multiple logistic regression and practice interpreting the parameters
* Introduce methods for assessing the accuracy of a classifier model

## Before class

* ISL ch 4.1-.3
* ESL ch 4.1, 4.4

## Slides and class materials

```{r slides}
knitr::include_url("/slides/logistic-regression.html", height = "505px")
```

* [Overview of logistic regression](/notes/logistic-regression/)

### Additional readings


## What you need to do

* Start [homework 02](/homework/core-regression/)
